Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-15 VegOzone.tex
Check abstract with grammarly

\affil[1]{Department of Geosciences, University of Oslo, Oslo, Norway}
\affil[2]{Department of Biosciences, University of Oslo, Oslo, Norway}
\affil[3]{Department of Environment and Geography, University of York, UK}
\affil[4]{Department of Geography, Ludwig Maximilians Universität, Munich, Germany}

@@

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies have been conducted to assess the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining essential climate variables, in particular ozone, over the past few decades. These data are used to assess the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and bright year (in particular in spring and early summer) with higher than average \chem{[O_3]} in April/May and a high number of episodic peak values in ozone volume mixing ratio (VMR) above $40\,\unit{ppb}$ in June--August, the latter in part attributable to forest fires in the Northern Hemisphere as well as warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that assessments that use generic parameterizations are likely to underestimate the risk of ozone damage in this region. We find that bespoke parameterizations of plant functional types (PFTs) for subarctic vegetation bio-types result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species this loss is up to $6\,\unit{\%}$ larger than projected from generic parameterizations. Efforts should be targeted towards accurately defining subarctic ecotypes' physiological response to essential climate variables to improve regional and global scale biogeochemical cycling under current and future climates.
\end{abstract}

>>>>

\affil[1]{Department of Geosciences, University of Oslo, Oslo, Norway}
\affil[2]{Department of Biosciences, University of Oslo, Oslo, Norway}
\affil[3]{Department of Environment and Geography, University of York, UK}
\affil[4]{Department of Geography, Ludwig Maximilian University of Munich, Munich, Germany}

@@

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations \chem{[O_3]}. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies assessed the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining essential climate variables, in particular ozone, over the past few decades. We evaluate the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and bright year, particularly in spring and early summer. Higher than average \chem{[O_3]} was observed in April/May while frequent episodes of ozone volume mixing ratios (VMRs) above $40\,\unit{ppb}$ occurred in June--August. These episodes are in part attributable to forest fires in the Northern Hemisphere and warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, environmental factors, and species-specific physiology. Our study suggests that using generic parameterizations in assessments likely leads to underestimating the risk of ozone damage in this region. We find that bespoke parameterizations of plant functional types (PFTs) for subarctic vegetation bio-types result in an ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. For some species, this loss is up to $6\,\unit{\%}$ larger than projected from generic parameterizations. Efforts should be targeted towards accurately defining subarctic vegetations' physiological response to essential climate variables. That could help to improve regional and global scale biogeochemical cycling under current and future climates.
\end{abstract}
